Oriented Gradient Histogram applied to P300 Detection

### 2017-12-04 article.tex
pequeño cambio

A local representation of the shape of the signal within the patch can be described by obtaining the gradient orientations on each of the $16$ blocks and creating a histogram of gradients.  This technique is based on Lowe's SIFT~\citep{Lowe2004} method, and it is biomimetically inspired in how the visual cortex detects shapes by analyzing orientations.   In order to calculate the histogram, the interval $[0-360]$ of possible angles is divided in $8$ bins, each one at $45$ degrees.
 For each spacial bin $ i,j = \{1,2,3,4\} $, corresponding to the indexes of each block $B_{i,j}$,  the orientations are accumulated in a  $3$-dimensional histogram $h$ through the following equation: %In each step, the histogram is computed by multiplying the length of the gradient vector by a Wang trilinear interpolation between the gradient angle and the bin, in order to give importance to the module of the gradient.

@@

Fig.~\ref{fig:sampledescriptor} shows an example of a patch and a scheme of the histogram computation. Fig.~\ref{patch} is a plot of the signal and the patch centered in the keypoint. In Fig.~\ref{histogrambines} the possible orientations on each patch are illustrated. They form the corresponding $\mathbf{kp}$-descriptor of $128$ coordinates. The first two blocks orientations are shown.  Following this procedure for every assigned keypoint, we obtain $N_{kp}$ descriptors.

\begin{figure}[H]
\centering
\subfigure[Plot of the signal, a keypoint and the corresponding patch.]{\includegraphics[width=.49\linewidth]{DiagramaConkeypoint.png}\label{patch}}
\subfigure[Orientations on each patch.]{\includegraphics[width=.49\linewidth]{histogramsbines.png}\label{histogrambines}}
%\includegraphics[width=9cm]{gradientszoom.png}
\caption{ Example of a patch and a scheme of the orientations histogram computation.}
\label{fig:sampledescriptor}
\end{figure}

@@

The aim is to identify the selected letter from the matrix. Previously, during training phase, descriptors from labeled segments are extracted.  The P300 templates are grouped in a template set called $ T $.  Rows are labeled 1-6, whereas columns are labeled 7-12.  The process has the following steps:

\begin{enumerate}
\item Highlight randomly the rows and columns from the matrix.  There are one row and one column that matches the selected letter by the user.
\item For $ k= {1-6} $: Obtain the descriptor from the image of the signal plot according to the method described in Section~\ref{SIFT}, $ d^r_1, \dots,  d^r_6 $ for rows and  $ d^c_1, \dots,  d^c_6 $. 
\item Compute

\begin{equation}
\hat{r} = \arg \min_u \sum_{q \in NN_T(d^r_u)}^{} \left\lVert q -  d^r_u \right\rVert ^2,\; u = 1,\dots,6
\label{eq:multiclassification}
\end{equation}

@@

\begin{equation}
\hat{c} = \arg \min_u \sum_{q \in NN_T(d^c_u)}^{} \left\lVert q -  d^c_u \right\rVert ^2,\; u = 7,\dots,12
\label{eq:multiclassification}
\end{equation}
where $q$ is a posible descriptor, $NN_T(d)$ is the set of descriptors nearest to $d$ and $\hat{r}$ and$$

@@

%The first step consists in extracting labeled descriptors from the training set for the  which are thus grouped together in two KD-tree \citep{Vedaldi2010} database structures. and group those marked as \textit{hit} in a template set $ T $

%BCI Simulation vs k-fold cross validation.  It is quite known and understood that because the same experience will not generate the same markers in the EEG, BCI Simulation is sometimes better suited to understand how the device will operate on a real online application (the graeat variability of BCI).

While doing classification based on local features, which encode partial information, one problem that frequently arises is how to ensemble the individual classification of each local characteristic to the whole image. %that were used to derive those descriptors.

The NBNN algorithm tackles this problem by comparing each image against a whole classification class. It is characterized by the set of descriptors closest to each one of the descriptors obtained for the query image.  The algorithm  find the predicted class $\hat{C}$ of a query image by means of the following minimization:

\begin{equation}
\hat{C} = \arg \min_C \sum_{}^{} \left\lVert d_i - NN_C(d_i) \right\rVert ^2
\label{eq:classification}
\end{equation}
where $C$ is the a class, $ d_i $ is a descriptor that belongs to the query image and $ NN_C(d_i) $ is the corresponding near neighbor descriptor for the class C.
%\noindent where the predicted class $\hat{C}$ of a query image is calculated as the class $ C $ that minimize the summation of the $ L^2 $ distance between each descriptor $ d_i $ that belongs to the query image and its corresponding near neighbour $ NN_C(d_i) $ descriptor for class C.

This basic method was adapted to the P300 Speller and the classifier works according to the following modified equation:

\begin{equation}
\hat{Q}_{r\,c} = \arg \min_q \sum_{x_i \epsilon NN_T^k(q)}^{} \left\lVert x_i - q \right\rVert ^2
\label{eq:multiclassification}
\end{equation}

The first step consists in extracting labeled \textit{hit} descriptors from the training set which are thus grouped together in a KD-tree \citep{Vedaldi2010} template set $ T $.  On decoding stage, 12 new images from the signal segments are generated, and their descriptors extracted.  The 12 descriptors are divided in rows and columns.  For each one of the 6 query descriptors $ q $ for row/column, its $ k $ nearest neighbours in the template set $T$, $ NN_T^k(q) $ (acquired during the training phase)  are obtained and the distance between each one of them, $ x_i $, and the query descriptor $ q $ is summarized.  The query descriptor $ q $ that minimizes this summation is the one that is chosen, first among the 6 rows, and then among the remaining 6 columns.  By performing this procedure, the correct letter can be identified by matching the row $ r $ and column $ c $ from the P300 speller matrix.

Especially in the case of the P300 response, the oddball paradigm requires that one of the stimuli need to be infrequent so that will unavoidably force the data to be unbalanced \citep{Tibon2015}.  The NBNN method suffers from biased classification on unbalanced classes \citep{Fornoni2014}.  By reversing the roles of the query and the class in Eq. \ref{eq:multiclassification}, it is only necessary to obtain the template set $ T $ with the learned descriptors representative of the P300 ERP, hence avoiding the problem of unbalanced classes.

%Additionally,

%El tema del balanceo lo mencionas y marcas lo cual está muy bien, pero no decí cómo lo solucionaste. Si llegaras a tener un poco de tiempo podrías mencionarlo.

>>>>

A local representation of the shape of the signal within the patch can be described by obtaining the gradient orientations on each of the $16$ blocks and creating a histogram of gradients.  This technique is based on Lowe's SIFT~\citep{Lowe2004} method, and it is biomimetically inspired in how the visual cortex detects shapes by analyzing orientations.   In order to calculate the histogram, the interval $[0-360]$ of possible angles is divided in $8$ bins, each one at $45$ degrees.

 For each spacial bin $ i,j = \{1,2,3,4\} $, corresponding to the indexes of each block $B_{i,j}$,  the orientations are accumulated in a  $3$-dimensional histogram $h$ through the following equation:

@@

Fig.~\ref{fig:sampledescriptor} shows an example of a patch and a scheme of the histogram computation. Fig.~\ref{patch} is a plot of the signal and the patch centered in the keypoint. In Fig.~\ref{histogrambines} the possible orientations on each patch are illustrated. They form the corresponding $\mathbf{kp}$-descriptor of $128$ coordinates. The first two blocks are shown.  Following this procedure for every assigned keypoint, we obtain $N_{kp}$ descriptors.

\begin{figure}[H]
\centering
\subfigure[Plot of the signal, a keypoint and the corresponding patch.]{\includegraphics[width=.49\linewidth]{DiagramaConkeypoint.png}\label{patch}}
\subfigure[Orientations on each patch.]{\includegraphics[width=.49\linewidth]{histogramsbines.png}\label{histogrambines}}
%\includegraphics[width=9cm]{gradientszoom.png}
\caption{ Example of a patch and a scheme of the orientation's histogram computation.}
\label{fig:sampledescriptor}
\end{figure}

@@

The aim is to identify the selected letter from the matrix. Previously, during training phase, descriptors from labeled segments are extracted.  These descriptors are the P300 templates which are grouped in a template set called $ T $.  Rows are labeled 1-6, whereas columns are labeled 7-12.  The process has the following steps:

\begin{enumerate}
\item Highlight randomly the rows and columns from the matrix.  There is one row and one column that should match the selected letter by the subject.
\item Obtain the descriptor from the image of the signal plot according to the method described in Section~\ref{SIFT}, $ d^r_1, \dots,  d^r_6 $ for rows and  $ d^c_7, \dots,  d^c_{12} $ for columns. 
\item Compute

\begin{equation}
\hat{r} = \arg \min_u \sum_{q \in NN_T(d^r_u)}^{} \left\lVert q -  d^r_u \right\rVert ^2,\; u = 1,\dots,6
\label{eq:multiclassificationrow}
\end{equation}

@@

\begin{equation}
\hat{c} = \arg \min_u \sum_{q \in NN_T(d^c_u)}^{} \left\lVert q -  d^c_u \right\rVert ^2,\; u = 7,\dots,12
\label{eq:multiclassificationcol}
\end{equation}

where $q$ is a template descriptor that belongs to the $k$ nearest neighbors  $NN_T(d^c_u)$.  By computing the aforementioned equations, the letter of the matrix can be determined from the intersection of the row $ \hat{r} $ and column $ \hat{c} $.

%The first step consists in extracting labeled \textit{hit} descriptors from the training set which are thus grouped together in a KD-tree \citep{Vedaldi2010} template set $ T $.  On decoding stage, 12 new images from the signal segments are generated, and their descriptors extracted.  The 12 descriptors are divided in rows and columns.  For each one of the 6 query descriptors $ q $ for row/column, its $ k $ nearest neighbours in the template set $T$, $ NN_T^k(q) $ (acquired during the training phase)  are obtained and the distance between each one of them, $ x_i $, and the query descriptor $ q $ is summarized.  The query descriptor $ q $ that minimizes this summation is the one that is chosen, first among the 6 rows, and then among the remaining 6 columns.  By performing this procedure, the correct letter can be identified by matching the row $ r $ and column $ c $ from the P300 speller matrix.

@@

\begin{figure}[H]
\centering
\includegraphics[width=10cm]{classification.png}
\caption{For each letter, single trial segments are averaged for the 6 rows and 6 columns.}
\label{fig:classification}
\end{figure}

Especially in the case of the P300 response, the oddball paradigm requires that one of the stimuli need to be infrequent so that will unavoidably force the data to be unbalanced \citep{Tibon2015}.  At the same time, the NBNN method suffers from biased classification on unbalanced classes \citep{Fornoni2014}.  By reversing the roles of the query and the class in equations \ref{eq:multiclassificationrow} and \ref{eq:multiclassificationcol}, it is only necessary to obtain the template set $ T $ with the learned descriptors representative of the P300 ERP, hence avoiding the problem of unbalanced classes. Figure \ref{fig:classification} schematize this basic algorithm where the classification is only performed against the template set T, for each one of the twelve classes, thus avoiding the unbalanced situation (2 hit vs. 10 no-hit).
